Abstract

### 2025-10-07 output/Main_text.Rmd
Explain new figure panels

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we must compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Hence, we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions
is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
In the first case, the two latent dimensions are independent,
with $`r TRANSFORM_MATRIX_TRANSP_INV`$ being an identity matrix.
In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
To keep the horizontal axis invariant,
we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
it only rotates the vertical axis [@harman_modern_1976]:

@@

The sample items have the same `r MODEL_ACRONYM` parameters
in both test spaces.
However, their multidimensional parameters differ due to the correlation among
the latent dimensions.
The parameters in both spaces,
along with their corresponding `r MODEL_ACRONYM` parameters,
are shown in Table \@ref(tab:example-items-table-out).

@@

Figure \@ref(fig:item-plot-out) illustrates the effect of
a positive correlation on the $`r MDISC_CORR_PARAM`$
(and consequently the $`r MIL_CORR_PARAM`$) value:
The components of the discrimination parameters tend to
*sum up* in the same direction as they get *more aligned*.
Thus, if the discrimination parameters have the same sign, as in items 1 and 5,
the $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
on the contrary, when the discrimination parameters have opposite signs
they tend to cancel each other out and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
as happens with items 2 and 3.
This effect can be especially noticed by comparing
the $`r MDISC_CORR_PARAM_ITEM`$ of items 3 and 5:
Their discrimination parameters are equal in absolute value,
so their $`r MDISC_CORR_PARAM_ITEM`$ values are equal in the orthogonal space;
however, the effect of the correlation shrinks the former and
stretches the latter.
Finally, note that item 4 has one discrimination parameter equal to 0,
so its $`r MDISC_CORR_PARAM_ITEM`$ is unaffected by the correlation.

@@

::: {custom-style="FigureNote"}
*Note.* The items are represented along with
a standard bivariate normal distribution, with null correlation (a),
or with correlation $`r CORR_OBL_OUT`$ (b-d).
The contour plots represent (from outer to inner)
the `r contour_breaks_out` of the maximum density.
:::

>>>>

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we must compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
Coordinates in a rectangular Cartesian system are commonly assumed
to be simply represented in the canonical basis.
To represent an item in the test space properly, however,
we must use its basis $`r TEST_SPACE_BASIS`$,
which implies pre-multiplying its coordinates
by $`r TRANSFORM_MATRIX_TRANSP_INV`$
to find the equivalent coordinates in the canonical basis.
Hence, we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions
is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
Their parameters are shown in Table \@ref(tab:example-items-table-out),
both in the original `r MODEL_ACRONYM` metric, and in the multidimensional
metric in two different cases, with different latent covariance structures.
In the first case (columns $`r CORR_ORTH_OUT`$),
the two latent dimensions are independent;
in the second one (columns $`r CORR_ORTH_OUT`$),
the correlation between the dimensions is .5.
Note that, although the `r MODEL_ACRONYM` parameters are the same in both cases,
their (correlation-based) multidimensional parameters change when different
correlations are used.

Panel (a) of Figure \@ref(fig:item-plot-out) shows
the item representation in the uncorrelated case,
while the correlated case is shown in panels (b) to (d)
under different conditions.
In all the four panels, the contours of a bivariate standard normal density
with the corresponding correlation are plot along with the items
(from outer to inner, the contours are the `r contour_breaks_out` of
the density at its mode).
Note that the stretched ellipses in panel (c) are the result of transforming
the distribution in the latent space to the test space,
according to \@ref(eq:trait-coords-spaces-transform).
In panel (b), the item coordinates are computed from
the agnostic version of the parameters.
These coordinates are the same ones as in the first case, but
their relative position with respect to the latent trait distribution is not
---This is the effect of disregarding the covariance structure.
Panel (c) shows the items with their coordinates computed using
the correlation-based version of the parameters.
To keep the horizontal axis invariant,
we have defined $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
it only rotates the vertical axis [@harman_modern_1976]:

@@

Comparing it to panel (b), we can see that the item coordinates are in general
not the same as the ones computed with the agnostic version of the parameters;
this effect is especially salient in items 2 and 3,
which are close to the 90% contour in the agnostic version,
but are almost along the 50% contour in the correlation-based one.
Finally, panel (d) illustrates the effect of plotting the coordinates directly
on the canonical basis; the relative positions and distances are correctly
represented (which allow a more direct comparison with panel (b)),
but the angles of the items with the axes are distorted:
For example, item 1 forms 45-degree angles with both axes,
instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).

@@

A comparison of panels (a) and (c) allows understanding the effect of
a positive correlation on the $`r MDISC_CORR_PARAM`$
(and consequently the $`r MIL_CORR_PARAM`$) value:
The components of the discrimination parameters tend to
*sum up* in the same direction as they get *more aligned*.
Thus, if the discrimination parameters have the same sign, as in items 1 and 5,
the $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
on the contrary, when the discrimination parameters have opposite signs
they tend to cancel each other out and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
as happens with items 2 and 3.
This effect can be especially noticed by comparing
the $`r MDISC_CORR_PARAM_ITEM`$ of items 3 and 5:
Their discrimination parameters are equal in absolute value,
so their $`r MDISC_CORR_PARAM_ITEM`$ values are equal in the orthogonal space;
however, the effect of the correlation shrinks the former and
stretches the latter.
Finally, note that item 4 has one discrimination parameter equal to 0,
so its $`r MDISC_CORR_PARAM_ITEM`$ is unaffected by the correlation.

@@

::: {custom-style="FigureNote"}
*Note.* The items are represented along with
a bivariate standard normal distribution, with null correlation (a),
or with correlation $`r CORR_OBL_OUT`$ (b-d).
The contour plots represent (from outer to inner)
the `r contour_breaks_out` of the maximum density.
:::
